Bayesian exponential random graph modeling of KWL Burial network

### 2020-11-19 analysis/paper/paper.Rmd
adjust network ties and work on result section

<!-- two types of network analysis -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches to characterize network properties at two distinct scales: node/edge level and graph level [@Peeples]. Node level focuses on the role of node in a network, such as centrality, representing the individual influence or social prominence in a group, while graph level assesses the whole network attributes, such as density, clustering in a network, to generalize relationship patterns [@Mills2017; @Peeples]. By quantifying those network properties, archaeologists can answer a wide range of research questions. Examples includes exploring the political centralization in the Kofun period in Japan through the hierarchical communication network constructed by prestige goods [@Mizoguchi2013], the investigation of long term inter-site relationships from the Epipalaeolithic to the early Neolithic in the Near East according to trade items [@Coward2013]. Regarding burials, @Sosna2013 examined spatial pattern of burials from the Early Bronze Age in Rebesovice with two hypothesized networks constructed according to cultural and chronological similarity between burials. Recently, complex network modeling evaluates networks at both node and graph level through simulations of particular processes and statistically testing the formation of network properties [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. Such application includes simulations and testing food exchange modes for Ancestral Pueblos on the aggregation of households in the American Southwest simulation [@Crabtree2015], or exploring the diffusion of fired bricks across Hellenistic Europe by comparing similarity networks of sites with random networks. Other examples show the assessment of hunter-gatherer exchange networks structure across the Kuril Islands using bootstrap simulation based on ceramic composition [@Erik2015]. This paper will use a novel Bayesian modeling approach to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organization.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are efficient computational tools for social network analysis because they incorporates prior information about the network configurations into the model and offers uncertainty quantification by evaluating the posterior distributions of the parameters associated with network configurations [@Caimo2015; @Nemmers2019]. <!-- add brief sentence to define posterior and prior to help the reader --> The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. MCMC avoids doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability estimates the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distribution for the priors is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of edge to low density for both network models. For covariates based on burial attributes, such as age, gender, and wealth index, we set the priors to follow a normal distribution with mean 0 and standard deviation 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 5)) to infer the kinship-based proximity. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for gwesp and gwedegree representing transitivity and popularity. We set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on the ritual (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the prior for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to illustrate an increased social inequality. This prior information derives from the theory about horizontal hierarchy, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for earlier network and N(-20, 6) for later network. It should be noted the idea of those settings is to reflect the relative differences between two networks, and the standard deviation was adjusted in terms of different networks to indicate uncertainty.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
To assess MCMC analysis, we first evaluate three diagnostics presented as visual summaries, including density plots, trace plots, and autocorrelation plots [@Hamra2013]. Those plots provide an informal way to diagnose model convergence by identifying any unexpected peaks or shapes in distribution. Then we access output from our two  models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network.

@@

<!--posterior stats for post-E and comparison-->
For the network after the arrival of Europeans, there are no significant effects of nodal and dyadic covariates, but we still have a clue of possible tendency. The nodal covariates of age-homophily, sex-homophily, ritual-homophily, and dyadic covariate of physical distance all show positive effects, while the burial value-homophily has a negative effect. The interesting thing is that the burials in the same wealth level may tend not to form relations. Similar to the network before the European arrival, the endogenous network variables, gwesp and gwdegree, all demonstrate positive effects significantly. However, there are differences in values of positive effects between two networks. The later network has a smaller gwesp effect and a larger gwedegree effect that may suggest a less tendency toward clustering but more tendency toward centralization after the European presence. Both posterior estimates presents symmetric distributions because of the posterior means are close to posterior medians (Table XX).

```{r vertex-bootstrap}
boostrap <- 1000
```
<!--comparison between two networks using bootstrap--> 
To understand whether the differences in posterior distributions is affected by different network size, we conducted sensitivity test using vertex bootstrap inference on networks that allows comparison for different small networks [@Chen2019]. Vertex bootstrap is a non-parametric method that conducts resampling of all vertices (i.e. node) of a network to quantify standard errors in network statistics estimation of interest. This enables the evaluation of uncertainty for networks and the comparison between two networks by examining the confidence interval for network population. We set the resampling size to 1,000 for network statistics, including density, clustering, and degree, for two network. The 95% confidence interval of each network statistic contains the observed network for both period, which indicate less sample size effects on our models.

<!--assessment of MCMC analysis and BGOF-->

```{r Bayesian-gof}
```

Figure shows the Bayesian goodness of fit diagnostics plots where the red lines represent the distributions of the observed data and the box plots represent the GOF distributions of 10000 network graphs simulated from ERGMs based on the estimated posterior distribution. The grey lines mark the 95% intervals. X-axis means the proportion of nodes

>>>>

<!-- two types of network analysis -->
Network analysis has been increasingly applied by archaeologists in recent years to deal with past interactions and explore the underlying mechanisms. There are two common approaches to characterize network properties at two distinct scales: node/edge level and graph level [@Peeples2019]. Node level focuses on the role of node in a network, such as centrality, representing the individual influence or social prominence in a group, while graph level assesses the whole network attributes, such as density, clustering in a network, to generalize relationship patterns [@Mills2017; @Peeples2019]. By quantifying those network properties, archaeologists can answer a wide range of research questions. Examples includes exploring the political centralization in the Kofun period in Japan through the hierarchical communication network constructed by prestige goods [@Mizoguchi2013], the investigation of long term inter-site relationships from the Epipalaeolithic to the early Neolithic in the Near East according to trade items [@Coward2013]. Regarding burials, @Sosna2013 examined spatial pattern of burials from the Early Bronze Age in Rebesovice with two hypothesized networks constructed according to cultural and chronological similarity between burials. Recently, complex network modeling evaluates networks at both node and graph level through simulations of particular processes and statistically testing the formation of network properties [@Brughmans2013; @Brughmans2018; @Freeman2004; @Salvini2010]. Such application includes simulations and testing food exchange modes for Ancestral Pueblos on the aggregation of households in the American Southwest simulation [@Crabtree2015], or exploring the diffusion of fired bricks across Hellenistic Europe by comparing similarity networks of sites with random networks. Other examples show the assessment of hunter-gatherer exchange networks structure across the Kuril Islands using bootstrap simulation based on ceramic composition [@Erik2015]. This paper will use a novel Bayesian modeling approach to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organization.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are efficient computational tools for social network analysis because they incorporates prior information about the network configurations into the model and offers uncertainty quantification by evaluating the posterior distributions of the parameters associated with network configurations [@Caimo2015; @Nemmers2019]. In Bayesian analysis, posterior distribution is an updated probability distribution after combining the prior derived from previous data or assumptions and our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. MCMC avoids doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporates our sample data and prior information through summary statistics from estimates of the ERGM parameters. Posterior probability estimates the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distribution for the priors is typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of edge to low density for both network models. For covariates based on burial attributes, such as age, gender, and wealth index, we specified a vague prior information that follows a normal distribution with mean at 0 and standard deviation at 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 5)) to infer the kinship-based proximity. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for gwesp and gwedegree representing transitivity and popularity. We set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on the ritual (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the prior for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to illustrate an increased social inequality. This prior information derives from the theory about horizontal hierarchy, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for earlier network and N(-20, 6) for later network. It should be noted the idea of those settings is to reflect the relative differences between two networks, and the standard deviation was adjusted in terms of different networks to indicate uncertainty.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
To assess MCMC analysis, we first evaluate three diagnostics presented as visual summaries, including density plots, trace plots, and autocorrelation plots [@Hamra2013]. Those plots provide an informal way to diagnose model convergence by identifying any unexpected peaks or shapes in distribution. Then we access output from our two  models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. Our Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. We set of 10000 network graphs simulated from the estimated posterior distribution in ERGMs. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network.

@@

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans, there are no significant effects of nodal and dyadic covariates, but we still have a clue of possible tendency. The nodal covariates of age-homophily, sex-homophily, ritual-homophily, and dyadic covariate of physical distance all show positive effects, while the burial value-homophily has a negative effect. The interesting thing is that the burials in the same wealth level may tend not to form relations. Similar to the network before the European arrival, the endogenous network variables, gwesp and gwdegree, all demonstrate positive effects significantly. However, there are differences in values of positive effects between two networks. The later network has a smaller gwesp effect and a larger gwedegree effect that may suggest a less tendency toward clustering but more tendency toward centralization after the European presence. Both posterior estimates presents symmetric distributions because of the posterior means are close to posterior medians (Table XX).

```{r vertex-bootstrap}
boostrap <- 1000
```

<!--comparison between two networks using bootstrap--> 
To understand whether the differences in posterior distributions are affected by different network sizes, we conducted a sensitivity test using vertex bootstrap inference on networks that allows a comparison for multiple networks of different sizes [@Chen2019]. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics estimation of interest. This enables the evaluation of uncertainty for networks and the comparison between networks by examining their confidence interval for the network population. We set the resampling size to 1,000 for network statistics, including density, clustering, and degree, for two networks. The result shows that the 95% confidence interval of each network statistic contains the observed network of both periods, which indicates our network is less sensitive to sample size. This suggests a limited sample size effect on both models. In addition, since the prior setting will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation.

```{r Bayesian-gof}
# the grey lines mark the 95% intervals. X-axis means the proportion of nodes
```

<!--assessment of MCMC analysis and BGOF--> 
The Bayesian goodness of fit diagnostics plots allows an evaluation of our model fit by comparing our observed data with the simulated distributions. We compared two models based on diagnostic plots that a good fit to estimated ERGM is indicated by observed data falls inside the 95% intervals.
